---
title: 'Coursera Reproducible Research Peer Assessment #2'
author: "Bill Cary"
date: "Sunday, December 14, 2014"
output: pdf_document
---

# NOAA Storm Database Exploration and Analysis
## Synopsis


## Data Analysis
### Set defaults
Set knitr to echo code by default
```{r setoptions, echo = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set echo=TRUE by default
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(knitr)
opts_chunk$set(echo = TRUE)

```

### Prepare the analysis environment
```{r prep}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Import required libraries
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
suppressMessages(library(ggplot2))     # General plotting functions
suppressMessages(library(plyr))        # Data manipulation
suppressMessages(library(dplyr))       # Data manipulation
suppressMessages(library(xtable))      # Table formatting in knitr document

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set paths for files and directory structure
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
path_data <- '../data/repdata-data-StormData.csv.bz2'

```

### Loading the data
Read the raw dataset into R.  Because the dataset is large and the initial load
is time consuming, the cache=TRUE option is used to decrease the runtime after
the initial execution of the analysis.
```{r loaddata, cache=TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Read the data into R
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
data <- read.csv(path_data)
```

### Explore the Data
For this exercise, the fields of interest are FATALITIES, INJURIES, PROPDMG,
PROPDMGEXP, CROPDMG and CROPDMGEXP.  The PROPDMGEXP and CROPDMGEXP fields
provide an "order of magnitude" describing the size of the monetary damages
(thousands, millions, billions, etc...)
```{r exploredata}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Generate summaries of the fields of interest
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
summary(data$FATALITIES)
summary(data$INJURIES)
summary(data$PROPDMG)
summary(data$PROPDMGEXP)
summary(data$CROPDMG)
summary(data$CROPDMGEXP)
```

From the summaries above, it does not appear that the FATALITIES, INJURIES,
PROPDMG and CROPDMG fields require cleansing/standardization.  However, the
PROPDMGEXP and CROPDMGEXP field contain significant duplication of values ("m"
and "M") that needs to be corrected before the data is useful for analysis.  In
addition, the exponents provided in those fields need to be applied to the
PROPDMG and CROPDMG values to put the values of each event on the same numeric
basis.

## Across the United States, which types of events (as indicated in the
## EVTYPE variable) are most harmful with respect to population health?
For the purposes of answering this question, I interpret the degree of harm of
an event with respect to population health as being the aggregate total of all
deaths and injuries (whether direct or indirect) caused by the event.  However,
I also make special note of the events which lead to the highest number of
direct and indirect deaths, as one can subjectively consider any injury, no
matter how severe, to be less severe than death.


#### Data Cleansing


## Across the United States, which types of events have the greatest economic
## consequences?

## Results


## Appendix

### Miscellaneous Support Information
Print out system and session information to support future readers in
replicating the environment used to produce the output, and in troubleshooting
any potential issues that arise when running the code.
```{r info}
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Print System and Session for reproducibility & troubleshooting.
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
print(sessionInfo())
# print(Sys.info())   omit Sys.info() for privacy reasons
```
